---
title: "Assignment 1"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    theme: "cosmo" # This theme has a slight purple tone. You can customize it further with CSS.
    highlight: tango
    css: styles.css  # Optional: You can add custom styles in an external CSS file
  pdf_document:
    toc: true
    toc_depth: 2
    latex_engine: xelatex
    includes:
      in_header: preamble.tex  # Optional: Add a LaTeX file to further control appearance
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(readxl)
```

## Guidelines

-   Use **R Markdown** to complete this assignment.\
-   Provide all your **code** to ensure that your work is **reproducible**.\
-   Include explanations or comments when necessary.

------------------------------------------------------------------------

# 1. Data Structures

### 1.1 Create a vector

Create a new chunk below this text. Then, create a vector called v1 with three elements, all of which are characters (enclosed in single ' ' or double " quotes). Save it and print the values.

```{r, include=FALSE}
# Create a character vector `v1`
v1 <- c("soybean", "maize", "alfalfa")
print(v1)
```

### 1.2 Convert vector to a factor

Create a vector v2 from v1, but convert it to a factor. Save and print the values. Change the order of the levels to any random order and print the values again.

```{r, include=FALSE}
v2 <- factor(v1)
print(v2)

# Change the order of the levels
v2 <- factor(v1, levels = c("soybean", "alfalfa", "maize"))
print(v2)
```

### 1.3 Create a list

Create a list l1 that contains v1, v2, and an additional object of class data.frame containing the following information:

Irrigation |  Value
-----------|-------
yes        |  200 
yes        |  200
no         |  0

```{r, include=FALSE}
# Create a data frame to include in the list
df_irr <- data.frame(irrigation = c("yes", "yes", "no"), value = c(200, 200, 0))

# Create a list `l1`
l1 <- list(v1, v2, df_irr)
print(l1)
```

------------------------------------------------------------------------

# 2. Reading and Saving Files, Accessing Elements

### 2.1 Read the tillage data

Read the file doy_weather.xlsx from the "data" folder into a new data frame df.

```{r, include=FALSE}
df <- read_xlsx("../../CiampittiLab_Rcourse_24/Data/doy_weather.xlsx")

print(df)
```

### 2.2 Check column types

Check the data types for all columns in the data frame. Some columns might be read as characters even though they represent numeric values. Convert any necessary columns to numeric.

```{r, include=FALSE}
# Check column types
print(str(df))

# Convert character columns to numeric
df <- df %>%
  mutate(across(c(TMIN, TMAX, RAIN), as.numeric))

print(str(df))
```

### 2.3 Replace missing values

Some columns contain missing values coded as -99. Replace all instances of -99 with NA in numeric columns only. Save the result in a new data frame called df_clean.

```{r, include=FALSE}
# Convert "-99" to NA in numeric columns using str_replace_all
df_clean <- df %>%
  mutate(across(where(is.numeric), ~str_replace_all(as.character(.), "-99", "NA") %>% as.numeric()))
```

### 2.3 Calculate mean Tmax

Calculate the mean of the TMAX column, ignoring missing values (NA). Save the result as mean_Tmax.

```{r, include=FALSE}
# Calculate mean Tmax
df_clean %>%
  summarise(mean_Tmax = mean(TMAX, na.rm = TRUE))
```

### 2.4 Compare temperatures

Create a new column Tmean, which is the average of TMAX and TMIN. Save the updated data frame as df_temp.

```{r, include=FALSE}
df_clean %>%
  mutate(Tmean = (TMAX + TMIN) / 2)
```

### 2.5 Day with the highest precipitation

Find the day with the highest precipitation (RAIN). Save the result as max_rain_day.

```{r, include=FALSE}
# Find the day with maximum precipitation
df_clean %>%
  filter(RAIN == max(RAIN, na.rm = TRUE)) %>%
  select(DOY, RAIN)
```

### 2.6 Subsetting data

Create a new data frame df_subset that includes only rows where precipitation (RAIN) is greater than 0.

```{r, include=FALSE}
# Subset data where RAIN > 0
df_clean %>%
  filter(RAIN > 0)
```

### 2.7 Sorting data

Sort the data frame df_clean by TMAX in descending order. Save the result as df_sorted.

```{r, include=FALSE}
# Sort by Tmax in descending order
df_sorted <- df_clean %>%
  arrange(desc(TMAX))
```

### 2.8 Summary statistics

Calculate summary statistics (mean, median, standard deviation) for the TMAX column. Save the result as df_summary.

```{r, include=FALSE}
# Summary statistics for Tmax
df_summary <- df_clean %>%
  summarise(
    mean_Tmax = mean(TMAX, na.rm = TRUE),
    median_Tmax = median(TMAX, na.rm = TRUE),
    sd_Tmax = sd(TMAX, na.rm = TRUE))

df_summary
```

### 2.9 Conditional Mutations with case_when

Create a new column Temperature_Category in df_clean, which classifies Tmean as:

"Low" if TMAX < 10  
"Moderate" if 10 <= TMAX <= 25  
"High" if TMAX > 25  

Save the updated data frame as `df_class` selecting DOY, TMAX, and Category columns.

```{r, include=FALSE}
# Add Temperature_Category using case_when
df_class <- df_clean %>%
  mutate(Category = case_when(
    TMAX < 10 ~ "Low",
    TMAX >= 10 & TMAX <= 25 ~ "Moderate",
    TMAX > 25 ~ "High",
    TRUE ~ NA_character_  # Catch any cases that don't match the conditions
  )) %>% 
  select(DOY, TMAX, Category)
```

