---
title: "Tidyverse"
author: "Luiz Felipe & Francisco Palmero"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# 1) Short review

-   Creating a chunk of code
-   Loading packages (tidyverse and readxl)
-   Creating an object
-   Commenting in .Rmd: in-chunk or out-of-chunk
-   Running a chunk of code
-   View an object in R studio
-   Pipe operator

# 2) Load packages

```{r}
library(tidyverse)
library(readxl)
```

# 3) Import data

```{r}
# Load weather data
df_weather <- read.csv("../Data/nfix_weather.csv") %>% select(-X)

# Load biomass data
df_biomass <- read.csv("../Data/nfix_biomass.csv") %>% select(-X) 

# Load yield data
df_yield <- read.csv("../Data/nfix_yield.csv") %>% select(-X)

str(df_weather)
str(df_biomass)
str(df_yield)
```

# 4) Tidyverse

-   What's tidy data?

### SELECT

Helper functions:
- starts_with()
- ends_with()
- contains()
- matches()
- num_range()
- grepl()
- between()
- across()

```{r}
# Select specific columns from the data
df_weather 

# Select columns whose names start with a given string
df_weather

# Select columns whose names end with a specific string.
df_weather

# Select columns whose names contain a specific substring.
df_weather
```

### ARRANGE

```{r}
# Reorder the data frame rows according to the lowest-to-highest yield
df_yield

# Reorder the data frame rows according to the highest-to-lowest yield
df_yield
```

### RENAME

```{r}
# Change column name
df_yield

# To upper or lower case (rename_with(tolower/touuper))
df_yield
df_yield
```

### FILTER

```{r}
# Filter rows
df_weather

# Filter rows where PP_B is less than or equal to 200
df_weather

# Filter rows where Yield_mgha is between 5 and 7 tons/ha (between() or using &)
df_yield
df_yield

# Filter KS1 sites (`%in%`)
df_weather

# Filter rows where Yield_mgha is not NA (!is.na())
df_yield

# Filter rows where Site contains "KS" (grepl)
df_weather
```

### MUTATE

```{r}
# Create a new column with total PP
df_weather

# Create a new column with total PP after a given column
df_weather

# Change type of data from character to factor (`mutate(across(c()))`)
df_biomass
```

### ROUND

```{r}
# Round yield values
df_yield
```

### SUMMARISE

```{r}
# Create a summary data frame with mean yield (na.rm=T)
df_yield
```

### GROUP_BY

```{r}
# Group by Treatment and summarize mean yield
df_yield

# Group by Site and Treatment and summarize mean yield
df_yield
```

### SEPARATE

```{r}
# Separate Site into Location and Year
df_yield
```

### UNITE

```{r}
# Create ID column (remove or not previous)
df_biomass
```

### DISTINCT

```{r}
# Get distinct Sites (distinct())
df_yield

# Get distinct combinations of Site and Treatment
df_yield
```

### COUNT

```{r}
# Count number of rows within each site (count())
df_yield
```


### CASE_WHEN

```{r}
# Create a new column that categorizes yields into "low", "medium", and "high"
# Yield_mgha < 3.5 ~ "low"; Yield_mgha >= 3.5 & Yield_mgha < 5 ~ "medium"; Yield_mgha >= 5 ~ "high"
df_yield

# Change only "Check" to "Control" (T~)
df_biomass

# Use base R to change the values of cells
#df_yield[1, 2] <- "Cheeck"
#df_yield[8, 2] <- "check"

# Solve the problem using %in%
# df_yield %>% 
#   mutate(Treatment = case_when(Treatment %in% c("Cheeck", "check") ~ "Check",
#                                TRUE ~ Treatment))
```

### PIPE

```{r}
# Chain multiple operations to process the data
df_yield %>%  
  filter(Treatment == "Check") %>% 
  group_by(Site) %>%
  mutate(mean_Yield_mgha = median(Yield_mgha)) %>% 
  summarise(mean_Yield_mgha) %>% 
  unique() %>% 
  separate(Site, into = c("Location", "Season"), sep = "_") %>% 
  rename(meanY = mean_Yield_mgha)
```

### STR_REPLACE

```{r}
# Replace a string in the Site column
df_yield %>% 
  mutate(across(Site, ~ str_replace(., pattern = "_", replacement = ":")))

# Let's create a non-desired pattern. Let the error be an underscore in the Check treatment.
# Solve the error so that you have the true name back 
df_yield %>% 
  mutate(Treatment = case_when(Treatment == "Check" ~ "C_heck",
                                TRUE ~ Treatment)) %>% 
  mutate(across(Treatment,
                ~ str_replace(., pattern = "_",
                              replacement = "")))
```

### IS.NA and NA_IF

```{r}
# Example Dataset: df_yield
df_yield_NA <- tibble(
  Site = c("KS1", "KS3", "KS4"),
  Treatment = c("Check", "N", "S"),
  Yield = c(50, 70, 65))

# 1. Use `na_if()` to replace the value "Check" in the Treatment column with NA
# df_yield_naif <- df_yield %>%
#   mutate(Treatment = na_if(Treatment, "Check"))
# 
# print(df_yield_naif)

# 2. Use `is.na()` to create a new column that flags where Treatment is NA
# df_yield_filtered <- df_yield_naif %>%
#   mutate(Flag_NA = is.na(Treatment))

# 3. Use `is.na()` to filter out rows where the Treatment is NA
# df_yield_filtered <- df_yield_naif %>%
#   filter(!is.na(Treatment))
```

### JOIN FUNCTIONS

```{r}
# Sample data frames
df_X <- tibble(
  Site_Season = c("IA1_2021", "IA2_2021", "IA3_2022"),
  Yield = c(150, 200, 180))

df_Y <- tibble(
  Site_Season = c("IA1_2021", "IA2_2021", "IA4_2022"),
  Biomass = c(50, 60, 55))

# INNER-JOIN: Includes common rows for x and y
df_X %>% 
  inner_join(df_Y, by = "Site_Season")

# LEFT_JOIN: Includes all rows from X and matched rows from Y
df_X %>% 
  left_join(df_Y, by = "Site_Season")

# RIGHT_JOIN: Includes all rows from Y and matched rows from X
df_X %>% 
  right_join(df_Y, by = "Site_Season")

# FULL_JOIN: Includes all rows from both X and Y
df_X %>% 
  full_join(df_Y, by = "Site_Season")
```

### PIVOT_WIDER

```{r}
# Pivot wider: Convert long data back to wide format
df_yield %>% 
  pivot_wider(names_from = Treatment, values_from = Yield_mgha)
```

### PIVOT LONGER

```{r}
# Pivot longer: Convert wide data to long format
df_weather %>% 
  dplyr::select(Site, season, PP_A, PP_B, PP_C, PP_D, PP_E) %>% 
  pivot_longer(cols = starts_with("PP"), names_to = "Period", values_to = "Precipitation") 
```

# 5) Final comments

-   Put all library() calls at the top of the script.
-   Organize your files into 'data', 'script/notebook'/'markdown', and 'output'.
-   Comment your code.
-   Object names: Don’t start with numbers & Keep them as short as understandable.
-   Each %>% should be followed by a new line.
-   Avoid long lines (start a new line after a comma).
-   Always use na.rm = TRUE in summarise() functions to handle NA values.
-   Use %>% for chaining functions to keep code clean and readable.
-   When joining datasets, ensure that the keys have consistent names and formats.







