---
title: "4 For Loops & Map & Function"
author: "Luiz Felipe & Francisco"
date: "`r Sys.Date()`"
output: html_document
---

**Today's Overview:**

-   Introduction to `for` loops in R
-   Using `map` functions from the `purrr` package as an alternative
-   Writing simple functions to use within loops

# Packages

```{r}
library(readxl)
library(tidyverse)
library(lubridate)
```

# Load data

```{r}
df_ks1 <- read_xlsx("../Data/Mult/weather_ks.xlsx", sheet = 1)
df_ks2 <- read_xlsx("../Data/Mult/weather_ks.xlsx", sheet = 2)

df_ne1 <- read_xlsx("../Data/Mult/weather_ne.xlsx", sheet = 1)
df_ne2 <- read_xlsx("../Data/Mult/weather_ne.xlsx", sheet = 2)

df_mi1 <- read_xlsx("../Data/Mult/weather_mi.xlsx", sheet = 1)
df_mi2 <- read_xlsx("../Data/Mult/weather_mi.xlsx", sheet = 2)

glimpse(df_ks1)
```

# What is a for loop?

-   Control flow statement for specifying iteration
-   Run a section of code repeatedly until a certain condition has been satisfied

## Basic structure

-   Header: defines the iteration

    -   Loop counter or loop variable

    -   Number of iterations is known before entering the loop

-   Body: code that is executed once per iteration

```{r}
# i is the loop variable that takes each value from 1 to 10
# print(i) is the code that runs each time the loop iterates

for (i in seq(1:10)) {
  print(i)
}

#Variable is a placeholder that takes on each value in the sequence
#Sequence is a collection of values (like numbers or items in a list) that you want to loop through
#The code inside the curly braces {} is what gets repeated
```

## Looping examples

### Example 1

```{r}
crop_vector <- c("soybean", "corn", "sorghum", "wheat")

for (crop in crop_vector) {
  print(crop)
}
```

### Example 2

```{r}
yields_ks <- c(4100, 5000, 5300, 2600, 3200)

for (i in 1:length(yields_ks)) {
  
  yield_kgha <- yields_ks[i] / 1000
  print(yield_kgha)

  }
```

### Example 3

```{r}
yields_kgha <- c(4100, 5000, 5300, 2600, 3200)

yields_ton <- c()  # Empty vector

for (i in 1:length(yields_kgha)) {
  
  yield_temp <- yields_kgha[i] / 1000  # Convert (save in a temporary object)
  yields_ton <- c(yields_ton, yield_temp)  # Append results
  
}

print(yields_ton) # Yield in ton/ha
```

### Example 4

```{r}
yields_df <- data.frame(
  Crop = c("soybean", "corn", "sorghum"),
  Yield_NorthFarm = c(4, 8, 6),
  Yield_Ashland = c(3, 7, 5))

print(yields_df)


# Empty object to store new data
total_yield_df <- NULL

# Loop over each row to calculate total yield for each crop
for (i in 1:nrow(yields_df)) {
  
  total_yield <- yields_df %>% 
    slice(i) %>%  
    mutate(Cumulative = Yield_NorthFarm + Yield_Ashland) %>%  
    select(Cumulative)  
  
  total_yield_df <- bind_rows(total_yield_df, total_yield)
}

print(total_yield_df)

yields_df <- bind_cols(yields_df, total_yield_df)

print(yields_df)
```

    
### Example 5

Problem: Wrangling across multiple excel files

```{r}
df_all_weather <- read_xlsx("../Data/Mult/weather_ks.xlsx", sheet = "ks_2013") %>%
  mutate(ID = "ks_2013") %>%
  rbind(
    read_xlsx("../Data/Mult/weather_mi.xlsx", sheet = "mi_2013") %>%
      mutate(ID = "mi_2013")
  ) %>%
  rbind(
    read_xlsx("../Data/Mult/weather_ne.xlsx", sheet = "ne_2013") %>%
      mutate(ID = "ne_2013")
  ) %>%
  rbind(
    read_xlsx("../Data/Mult/weather_ks.xlsx", sheet = "ks_2020") %>%
      mutate(ID = "ks_2020")
  ) %>%
  rbind(
    read_xlsx("../Data/Mult/weather_mi.xlsx", sheet = "mi_2020") %>%
      mutate(ID = "mi_2020")
  ) %>%
  rbind(
    read_xlsx("../Data/Mult/weather_ne.xlsx", sheet = "ne_2020") %>%
      mutate(ID = "ne_2020"))

df_all_weather_clean <- df_all_weather %>%
  dplyr::select(-c(DEWP:RHUM)) %>% 
  mutate(DATE = as.Date(as.character(DATE), format = "%Y%j"),  
         across(where(is.numeric), ~ na_if(., -99))) %>%
  group_by(ID) %>%  
  mutate(AMP = TMAX - TMIN,
         Avg_AMP = mean(AMP, na.rm = TRUE))
```

Let's try some automation

```{r}
file_paths_test <- list.files(path = "../Data/Mult/", pattern = "weather", full.names = TRUE)

sheet_names_test <- excel_sheets(file_paths[1])

read_xlsx(file_paths_test[1], sheet = sheet_names_test[1]) %>%
  dplyr::select(-c(DEWP:RHUM)) %>% 
  mutate(DATE = as.Date(as.character(DATE), format = "%Y%j"),  
         across(where(is.numeric), ~ na_if(., -99))) %>%
  mutate(AMP = TMAX - TMIN,
         Avg_AMP = mean(AMP, na.rm = TRUE))
```

Let's loop it!!

```{r}
file_paths <- list.files(path = "../Data/Mult/", pattern = "weather", full.names = TRUE)

weather_dfs <- data.frame(NULL)

for (file in seq_along(file_paths)) {
  
  sheet_names <- excel_sheets(file_paths[file])
  
  for (sheet in sheet_names) {
    
    temp_df <- read_xlsx(file_paths[file], sheet = sheet) %>%
      dplyr::select(-c(DEWP:RHUM)) %>%                      
      mutate(
        DATE = as.Date(as.character(DATE), format = "%Y%j"),  
        across(where(is.numeric), ~ na_if(., -99))) %>%
      mutate(
        AMP = TMAX - TMIN,                                      
        Avg_AMP = mean(AMP, na.rm = TRUE)) %>%
      mutate(ID = sheet)                                 
    
    weather_dfs <- rbind(weather_dfs, temp_df)
  }
}

combined_weather_df <- bind_rows(weather_dfs)

print(combined_weather_df)
```

    


    
    
    
    
    
