---
title: "4 For Loops & Map & Function"
author: "Luiz Felipe & Francisco"
date: "`r Sys.Date()`"
output: html_document
---

**Today's Overview:**

-   Introduction to `for` loops in R
-   Writing simple functions to use within loops
-   Using `map` functions from the `purrr` package as an alternative


# 0. Packages

```{r}
library(readxl)
library(tidyverse)
library(lubridate)
```

# 1. Load data

```{r}
df_ks1 <- read_xlsx("../Data/Mult/weather_ks.xlsx", sheet = 1)
df_ks2 <- read_xlsx("../Data/Mult/weather_ks.xlsx", sheet = 2)

df_ne1 <- read_xlsx("../Data/Mult/weather_ne.xlsx", sheet = 1)
df_ne2 <- read_xlsx("../Data/Mult/weather_ne.xlsx", sheet = 2)

df_mi1 <- read_xlsx("../Data/Mult/weather_mi.xlsx", sheet = 1)
df_mi2 <- read_xlsx("../Data/Mult/weather_mi.xlsx", sheet = 2)

glimpse(df_ks1)
```

# 2. For loop

What is a for-loop / for loop?

-   Control flow statement for specifying iteration
-   Run a section of code repeatedly until a certain condition has been satisfied

## Basic structure

-   Header: defines the iteration

    -   Loop counter or loop variable

    -   Number of iterations is known before entering the loop

-   Body: code that is executed once per iteration

```{r}
# i is the loop variable that takes each value from 1 to 10
# print(i) is the code that runs each time the loop iterates

for (i in seq(1:10)) {
  print(i)
}

for (placeholder in seq(1:5)) {
  
  print(placeholder)

  }

# Variable is a placeholder that takes on each value in the sequence
# Sequence is a collection of values (like numbers or items in a list) that you want to loop through
# The code inside the curly braces {} is what gets repeated
```

### Example 1

```{r}
crop_vector <- c("soybean", "corn", "sorghum", "wheat")

for (i in 1:length(crop_vector)) {
  print(crop_vector[i])
}

for (i in crop_vector) {
  print(i)
}

```

### Example 2

```{r}
yields_ks <- c(4100, 5000, 5300, 2600, 3200)

for (i in 1:length(yields_ks)) {
  
  yield_kgha <- yields_ks[i] / 1000
  print(yield_kgha)
  
  }
```

### Example 3

```{r}
yields_kgha <- c(4100, 5000, 5300, 2600, 3200)

yields_ton <- c(NULL)  # Empty vector

for (i in 1:length(yields_kgha)) {
  
  yield_temp <- yields_kgha[i] / 1000  # Convert (save in a temporary object)
  yields_ton <- c(yields_ton, yield_temp)  # Append results
  # print(yields_ton)
  
}

print(yields_ton) # Yield in ton/ha
```

### Example 4

```{r}
yields_df <- data.frame(
  Crop = c("soybean", "corn", "sorghum"),
  Yield_NorthFarm = c(4, 8, 6),
  Yield_Ashland = c(3, 7, 5))

print(yields_df)


# Empty object to store new data
total_yield_df <- NULL

# Loop over each row to calculate total yield for each crop
for (i in 1:nrow(yields_df)) {
  
  total_yield <- yields_df %>% 
    slice(i) %>%  
    mutate(Cumulative = Yield_NorthFarm + Yield_Ashland) %>%  
    select(Cumulative)  
  
  total_yield_df <- bind_rows(total_yield_df, total_yield)
}

print(total_yield_df)

yields_df <- bind_cols(yields_df, total_yield_df)

print(yields_df)
```

### Example 5

Problem: Wrangling across multiple excel files

```{r}
df_all_weather <- read_xlsx("../Data/Mult/weather_ks.xlsx", sheet = "ks_2013") %>%
  mutate(ID = "ks_2013") %>%
  rbind(
    read_xlsx("../Data/Mult/weather_mi.xlsx", sheet = "mi_2013") %>%
      mutate(ID = "mi_2013")
  ) %>%
  rbind(
    read_xlsx("../Data/Mult/weather_ne.xlsx", sheet = "ne_2013") %>%
      mutate(ID = "ne_2013")
  ) %>%
  rbind(
    read_xlsx("../Data/Mult/weather_ks.xlsx", sheet = "ks_2020") %>%
      mutate(ID = "ks_2020")
  ) %>%
  rbind(
    read_xlsx("../Data/Mult/weather_mi.xlsx", sheet = "mi_2020") %>%
      mutate(ID = "mi_2020")
  ) %>%
  rbind(
    read_xlsx("../Data/Mult/weather_ne.xlsx", sheet = "ne_2020") %>%
      mutate(ID = "ne_2020"))

df_all_weather_clean <- df_all_weather %>%
  dplyr::select(-c(DEWP:RHUM)) %>% 
  mutate(DATE = as.Date(as.character(DATE), format = "%Y%j"),  
         across(where(is.numeric), ~ na_if(., -99))) %>%
  group_by(ID) %>%  
  mutate(AMP = TMAX - TMIN,
         Avg_AMP = mean(AMP, na.rm = TRUE))
```

Let's try some automation

```{r}
file_paths_test <- list.files(path = "../Data/Mult/", pattern = "weather", full.names = TRUE)

sheet_names_test <- excel_sheets(file_paths_test[1])

read_xlsx(file_paths_test[1], sheet = sheet_names_test[2]) %>%
  dplyr::select(-c(DEWP:RHUM)) %>% 
  mutate(DATE = as.Date(as.character(DATE), format = "%Y%j"),  
         across(where(is.numeric), ~ na_if(., -99))) %>%
  mutate(AMP = TMAX - TMIN,
         Avg_AMP = mean(AMP, na.rm = TRUE))
```

Let's loop it!!

```{r}
file_paths <- list.files(path = "../Data/Mult/", pattern = "weather", full.names = TRUE)

weather_dfs <- data.frame(NULL)

for (file in seq_along(file_paths)) {
  
  sheet_names <- excel_sheets(file_paths[file])
  
   for (sheet in sheet_names) {
     
     temp_df <- read_xlsx(file_paths[file], sheet = sheet) %>%
       dplyr::select(-c(DEWP:RHUM)) %>%                      
       mutate(
         DATE = as.Date(as.character(DATE), format = "%Y%j"),  
         across(where(is.numeric), ~ na_if(., -99))) %>%
       mutate(
         AMP = TMAX - TMIN,                                      
         Avg_AMP = mean(AMP, na.rm = TRUE)) %>%
       mutate(ID = sheet)                                 
    
   weather_dfs <- rbind(weather_dfs, temp_df)
  }
}

combined_weather_df <- bind_rows(weather_dfs)

print(combined_weather_df)
```

# 2. Custom functions

-   "Recipe" (inputs/process/output)
-   Avoid repetition, cleaner code, reproducible

## Basic structure

-   Function name
-   Arguments: inputs
-   Code block `{ }`: steps the function should perform
-   Return value: output

```{r}
my_function <- function(argument1, argument2) { 
  result <- argument1 * argument2
  return(result)
}
```

### Example 1: 

```{r}
my_multiply <- function(value1, value2) {
  multiplication <- value1 * value2
  return(multiplication)
}

my_multiply(value1 = 209470927, value2 = 0.00001)
```

### Example 2

```{r}
clean_weather_xlsx <- function(file_path, sheet) {
  
  df <- read_excel(path = file_path, sheet = sheet)
  
  df_clean <- df %>%
    dplyr::select(-c(DEWP:RHUM)) %>% 
    mutate(DATE = as.Date(as.character(DATE), format = "%Y%j"),  
           across(where(is.numeric), ~ na_if(., -99))) %>%
    mutate(AMP = TMAX - TMIN,              
           Avg_AMP = mean(AMP, na.rm = TRUE)) 
  
  return(df_clean)
}

clean_weather_xlsx(file_path = "../Data/Mult/weather_ks.xlsx", sheet = 1)
```

### Example 3

```{r}
plot_vars_vs_doy <- function(object_df) {
  
  weather_vars <- c("TMAX", "TMIN", "RAIN", "SRAD", "AMP")
  
  plots <- list()
  
  for (var in weather_vars) {
    
    date_to_doy <- object_df %>% mutate(DOY = yday(DATE))
    
    p <- ggplot(date_to_doy, aes(x = DOY, y = .data[[var]])) + 
      geom_line() +
      labs(title = paste(var, "vs DOY"), x = "Day of year", y = var) +
      theme_bw()
    
    plots[[var]] <- p
  }
  
  return(plots)
}

plot_vars_vs_doy(object_df = df_clean)
```

# 3. Map

-   `purrr` package
-   More concise than for loops and integrate better with the `tidyverse` workflow

## Basic structure

-   `map(.x = list/vector/df, .f = function)`

    -   .x is the list or vector you want to iterate over

    -   .f is the function you want to apply to each element of .x

    -   map() returns a list

    -   Use a dot "." as a placeholder for the left-hand side of the pipe
    
    -   The utility of map with nested data

### Example 1

```{r}
crop_vector <- c("soybean", "corn", "sorghum", "wheat")

# Using map to print each crop
map(crop_vector, ~ print(.x))

# Using the pipe with map
crop_vector %>% 
  map(~ print(.x))
```

-   Map functions truly shine when dealing with nested data 
-   Nested data refers to data structures within structures, such as lists within lists or grouped data frames

### Example 2

```{r}
df_nested <- combined_weather_df %>% 
  group_by(ID) %>% 
  nest()

View(df_nested)
  
df_nested %>% 
  mutate(n_days = map(data, ~ nrow(.)) %>% as.numeric(.),
         n_cols = map(data, ~ ncol(.)) %>% as.numeric(.)) %>% 
  unnest()
```

### Example 3

```{r}
df_nested_tmax <- df_nested %>% 
  mutate(
    tmax_plot = map(data, ~ ggplot(.x, aes(x = yday(DATE), y = TMAX)) +
                           geom_line() +
                           labs(x = "Day of year", y = "TMAX") +
                           theme_minimal()))

print(df_nested_tmax$tmax_plot[1])
```

### Example 4

```{r}
df_nested_map2 <- df_nested %>% 
  mutate(
    tmax_plot = map2(data, ID, ~ ggplot(.x, aes(x = yday(DATE), y = TMAX)) +
                           geom_line() +
                           labs(
                             title = paste("TMAX vs DOY for", .y),  # Add ID to the title
                             x = "DOY", 
                             y = "TMAX") +
                           theme_bw()))

print(df_nested_map2$tmax_plot[[1]])
df_nested_map2 %>% pull(tmax_plot)
```

### Example 5

```{r}
crop_data <- data.frame(
  Crop = rep(c("Soybean", "Corn", "Wheat"), each = 5),
  Year = rep(2015:2019, times = 3),
  Yield = c(3.1, 3.3, 3.5, 3.4, 3.2, 8.1, 7.9, 8.3, 8.2, 8.0, 5.1, 5.3, 5.2, 5.4, 5.3))

# Nest the data by Crop
nested_crop_data <- crop_data %>%
  group_by(Crop) %>%
  nest()

View(nested_crop_data)

# View(data.frame)

nested_crop_data <- nested_crop_data %>%
  mutate(stats = pmap(.l = list(data, Crop),
    ~ data.frame(
      crop = ..2,
      mean_Yield = mean(..1$Yield),  
      sd_Yield = sd(..1$Yield), 
      max_Yield = max(..1$Yield),
      min_Yield = min(..1$Yield)
    )
  ))

unnested_data <- nested_crop_data %>%
  select(Crop, stats) %>%
  unnest(cols = stats, names_sep = "_")

print(unnested_data)
```
























