
# Packages
```{r}
library(tidyverse)
library(emmeans)
library(car)
library(lmerTest)
```


# 1) RCBD with One Factor: One-Way ANOVA

## 1.1) Looking the data
```{r}
df <- read.csv("../Data/Nfer_Corn.csv")

summary(df)
glimpse(df)
```


### 1.1.1) Plotting the data
```{r}
df %>% 
  ggplot(aes(Nrate_kg.ha,GY_kg.ha)) +
  geom_point(aes(color=as.character(block))) +
  scale_x_continuous(limits = c(0,380))+
  scale_y_continuous(limits = c(0,16000)) +
  geom_hline(yintercept = mean(df$GY_kg.ha))
```


## 1.2) Transforming the data for conducting an ANOVA
```{r}
df <- df %>% 
  mutate(across(c(block,Nrate_kg.ha),as.factor))
```


## 1.3) Fitting the model

$$y_{ij} = \mu + \tau_i + \epsilon_{ij}$$
for $i = 1,2,...,6$ and $j=1,2,...,24$


$y_{ij}$ is the response of the *jth* experimental unit given treatment the *ith* N rate

$\mu$ is some fixed constant

$\tau_i$ is the effect of the *ith* N rate as deviation from $\mu$

$\epsilon_{ik}$ is the experimental error


### 1.3.1) No restriction over the $\tau$

```{r}
m1 <- lm(GY_kg.ha~Nrate_kg.ha+block, data=df)
summary(m1)
```



### 1.3.2) Restriction over the $\tau$

$$\sum_{i=1}^t \tau_i = 0$$

```{r}
contrasts(df$Nrate_kg.ha) <- contr.sum(6)

m2 <- lm(GY_kg.ha~Nrate_kg.ha+block, data=df)
summary(m2)

```



## 1.4) Perfoming the ANOVA
```{r}
anova(m1)
anova(m2)
```


## 1.5) Another way of conducting the ANOVA
```{r}
m2 <- aov(GY_kg.ha~Nrate_kg.ha+block, data=df)
summary(m2)
```


## 1.6) Checking assumptions
```{r}
res <- residuals(m1)

# Normality
shapiro.test(res) # H0: residuals are normally distributed; do not reject H0 if p-value > 0.05

qqnorm(res);qqline(res)

# Homogeinity of variance
plot(fitted(m1), res);abline(0,0)

car::leveneTest(GY_kg.ha~Nrate_kg.ha, 
           data = df)

```



## 1.7) Pairwise comparisons

Obtain the estimated means
```{r}
m1_means <- emmeans(m1,~Nrate_kg.ha)
m1_means
```


### 1.7.1) t-tests
```{r}
# p-values for the pairwise comparisons
graphics::pairs(m1_means,
                level = 0.95)

# Confidence intervals for the pairwise comparions
confint(graphics::pairs(m1_means,
              adjust="none",
              level = 0.95))

# Obteining letters for showing differences
multcomp::cld(m1_means,adjust = "none", 
              Letters = letters,
              level = 0.95,
              decreasing = TRUE)
```


### 1.7.2) Tukey's HSD
```{r}
# p-values for the pairwise comparisons
graphics::pairs(m1_means,
                level = 0.95)

# Confidence intervals for the pairwise comparions
confint(graphics::pairs(m1_means,
              adjust="tukey",
              level = 0.95))

# Obteining letters for showing differences
multcomp::cld(m1_means,adjust = "sidak", 
              Letters = letters,
              level = 0.95,
              decreasing = TRUE)
```






# 2) Split-Plot in a RCBD: ANOVA

Plot -> Density
Subplot -> Hybrid
y -> seed weight $g~plant^{-1}$

## 2.1) Looking the data
```{r}
df2 <- read.csv("../Data/Sorgh_Dens_Hyb.csv") %>% 
  mutate(across(c(Block,Density,Hyb), as.factor))

summary(df2)
glimpse(df2)
```


## 2.2) Fitting the model


$$y_{ijk} = \mu + \alpha_j + \rho_i + \delta_{ij} + \beta_k + (\alpha\beta)_{jk} + \epsilon_{ijk},$$

$$\delta_{ij} \overset{\text{i.i.d.}}{\sim}  N(0,\sigma^2_w), $$


$$\epsilon_{ibk} \overset{\text{i.i.d.}}{\sim}  N(0,\sigma^2_{\epsilon}).$$

Furthermore $\delta_{ij}$ and $\epsilon_{ibk}$ are independent.

$y_{ijk}$ is the response corresponding to the ith block, the j level of factor Density, the k level of factor Hybrid.

$\mu$ is some fixed constant

$\alpha_j$ is the effect of the j level of Density

$\rho_i$ is the effect of the ith block

$\delta_{ij}$ is the whole-plot random error

$\beta_k$ is the effect of the k level of Hybrid

$(\alpha\beta)_{jk}$ is the interaction effect of level j of Density and level k of Hybrid

$\epsilon_{ijk}$ is the subplot random error




### 2.2.1) Using base R
```{r}
m3 <- aov(Weight~Density+Error(Block)+Hyb+Density:Hyb,
    data = df2)

```

### 2.2.2) Using lmer function from lmerTest package
```{r}
m4 <- lmerTest::lmer(Weight~Density*Hyb+(1|Density:Block),
               data=df2)

summary(m4)
```


## 2.3) Conducting ANOVA
```{r}
# Type I sum of squares
summary(m3)
# Type III sum of squares
anova(m4)
```


## 2.4) Checking assumptions
```{r}
res <- residuals(m4)

# Normality
shapiro.test(res) # H0: residuals are normally distributed; do not reject H0 if p-value > 0.05

qqnorm(res);qqline(res)

# Homogeinity of variance
plot(fitted(m4), res);abline(0,0)

car::leveneTest(Weight~Density*Hyb, 
           data = df2)

```


## 2.5) Pairwise comparisons

There was no significant interaction, therefore, we can compare the main effects directly


Obtain the estimated means for Density main effect
```{r}
m4_Dens_mean <- emmeans(m4,~Density)
m4_Dens_mean
```


Obtain the estimated means for Hybrid main effect
```{r}
m4_Hyb_mean <- emmeans(m4,~Hyb)
m4_Hyb_mean
```



### 2.5.1) Density Comparisons: Tukey's HSD
```{r}
# p-values for the pairwise comparisons
graphics::pairs(m4_Dens_mean,
                level = 0.95)

# Confidence intervals for the pairwise comparisons
confint(graphics::pairs(m4_Dens_mean,
              adjust="tukey",
              level = 0.95))

# Obtaining letters for showing differences
multcomp::cld(m4_Dens_mean,adjust = "sidak", 
              Letters = letters,
              level = 0.95,
              decreasing = TRUE)
```



### 2.5.2) Hybrids Comparisons: Tukey's HSD
```{r}
# p-values for the pairwise comparisons
graphics::pairs(m4_Hyb_mean,
                level = 0.95)

# Confidence intervals for the pairwise comparisons
confint(graphics::pairs(m4_Hyb_mean,
              adjust="tukey",
              level = 0.95))

# Obtaining letters for showing differences
multcomp::cld(m4_Hyb_mean,adjust = "sidak", 
              Letters = letters,
              level = 0.95,
              decreasing = TRUE)
```







