
# Packages
```{r message=FALSE}
library(tidyverse)
library(lme4)
library(nlme)
```


# 1) RCBD with One Factor: One-Way ANOVA

## 1.1) Looking the data
```{r}
df <- read.csv("../Data/Nfer_Corn.csv")
df <- read.csv("../CiampittiLab_Rcourse_24/Data/Nfer_Corn.csv")

summary(df)
glimpse(df)
```


### 1.1.1) Plotting the data
```{r}
df %>% 
  ggplot(aes(Nrate_kg.ha,GY_kg.ha)) +
  geom_point(aes(color=as.character(block))) +
  scale_x_continuous(limits = c(0,380))+
  scale_y_continuous(limits = c(0,16000)) +
  geom_hline(yintercept = mean(df$GY_kg.ha))
```


## 1.2) Transforming the data for conducting an ANOVA
```{r}
df <- df %>% 
  mutate(across(c(block,Nrate_kg.ha),as.factor))
```



## 1.3) Fitting the model: Random Effects Model

$$y_{ij} = \mu + \upsilon_i + \epsilon_{ij}$$
for $i = 1,2,...,6$ and $j=1,2,...,24$

$$\upsilon_i \sim N(0, \sigma^2_{\upsilon})$$



$$\epsilon_{ij} \sim N(0, \sigma^2_{\epsilon})$$

$y_{ij}$ is the response of the *jth* observation given treatment the *ith* N rate

$\mu$ is some fixed constant

$\upsilon_i$ effect of the ith randomly selected N rate.

$\epsilon_{ij}$ is the experimental error


```{r}
m1_random <- lmer(GY_kg.ha ~ (1|Nrate_kg.ha),
     data = df)

summary(m1_random)
```


```{r}
m1_random@pp
```


```{r}
m1_rand_nlme <- nlme::lme(GY_kg.ha ~ 1, random = ~1|Nrate_kg.ha,
          data = df)

summary(m1_rand_nlme)
```


Intraclass Correlation
```{r}
5554430/(5554430+1290905)
```


Best Linear Unbiased Predictor (BLUP)

$$\hat\beta_0 + \hat\upsilon_i$$

$$\hat\upsilon_i = \frac{\sigma^2_{\upsilon}}{\sigma^2_{\upsilon} + \sigma^2_{\epsilon} / n} \left( \bar{y}_i - \bar{y} \right)$$


Getting the BLUP for N0
```{r}
N0_mean = as.numeric(df %>% 
  filter(Nrate_kg.ha == "0") %>% 
  summarise(N0_mean = mean(GY_kg.ha)))

# Getting \hat\upsilon_i
hat_ups = (5554430/(5554430+1290905/4))*(N0_mean - mean(df$GY_kg.ha))
hat_ups

# Getting the BLUP for N0
10102.5+hat_ups
```


$\hat\upsilon_i$ and BLUP using lme4 and base R
```{r}
# Getting upsilon hat
ranef(m1_random)

# Getting the BLUPs
predict(m1_random)
```



### 1.3.1) Random Effects with N Rate and Blocks

$$y_{ij} = \mu + \upsilon_i + \gamma_{ij} + \epsilon_{ij}$$

for $i = 1,2,...,6$ and $j=1,2,...,24$

$$\upsilon_i \sim N(0, \sigma^2_{\upsilon})$$

$$\gamma_{ij} \sim N(0, \sigma^2_{\gamma})$$


$$\epsilon_{ik} \sim N(0, \sigma^2_{\epsilon})$$


$y_{ij}$ is the response of the *jth* observation given treatment the *ith* N rate

$\mu$ is some fixed constant

$\upsilon_i$ effect of the ith randomly selected N rate (captures variability among N rates, N effect over plant physiology).

$\gamma_{ij}$ effect of the ith randomly selected blocks (captures variability among blocks, e.g. soil variability).

$\epsilon_{ik}$ is the experimental error


```{r}
m2_random <- lmer(GY_kg.ha ~ (1|Nrate_kg.ha) + (1|block),
     data = df)

summary(m2_random)
```


Random Effects and BLUPs
```{r}
# Getting upsilon hat
ranef(m2_random)

# Getting the BLUPs
predict(m2_random)
```


Intraclass Correlation for N Rates within Blocks
```{r}
(5642072+350545)/(5642072+350545+940360)
```



## 1.4) Fitting the model: Mixed Effects Model with Blocks as random

$$y_{ij} = \mu + \tau_i + \upsilon_j + \epsilon_{ij}$$

for $i = 1,2,...,6$ and $j=1,2,...,24$


$$\upsilon_j \sim N(0, \sigma^2_{\upsilon})$$


$$\epsilon_{ij} \sim N(0, \sigma^2_{\epsilon})$$



$y_{ij}$ is the response of the *jth* experimental unit given treatment the *ith* N rate

$\mu$ is some fixed constant

$\tau_i$ is the effect of the *ith* N rate as deviation from $\mu$

$\upsilon_j$ the effect of the jth randomly selected block

$\epsilon_{ij}$ is the experimental error


```{r}
m1_mixed <- lmer(GY_kg.ha ~ Nrate_kg.ha + (1|block),
     data = df)

summary(m1_mixed)
```



Getting the $\hat\upsilon_i's$

```{r}
ranef(m1_mixed)
```


Getting the BLUPs
```{r}
predict(m1_mixed)
```



### 1.4.1) Looking at fixed blocks

$$y_{ij} = \mu + \tau_i + \beta_j + \epsilon_{ij}$$

for $i = 1,2,...,6$ and $j=1,2,...,24$


$y_{ij}$ is the response of the *jth* experimental unit given treatment the *ith* N rate

$\mu$ is some fixed constant

$\tau_i$ is the effect of the *ith* N rate as deviation from $\mu$

$\beta_j$ the effect of the jth block

$\epsilon_{ij}$ is the experimental error


```{r}
m1_fixed <- lm(GY_kg.ha ~ Nrate_kg.ha + block,
     data = df)

summary(m1_fixed)
```


### 1.4.2) ANOVA
```{r}
car::Anova(m1_mixed, type="II")
car::Anova(m1_fixed, type="II")
```




# 2) Linear Mixed Model

The data we will use in this example was adapted from Fernandez et al. (2022): https://academic.oup.com/jxb/article/73/11/3597/6547901#357375884

```{r}
df_corn <- read.csv("../Data/Corn_KN_KW_Yield.csv") %>% 
  mutate(across(c(StudyId,WaterCond), as.factor),
         Yield = Yield*1000)

df_corn
```


```{r}
df_corn %>% 
  group_by(StudyId) %>% 
  count(YOR)
  distinct(YOR)
```


## 2.1) Explore the data visually

Define the colors
```{r}
fill_color <- scale_fill_manual(values = c("1"="#922b21","2"="#76448a","3"="#2471a3","4"="#17a589","5"="#cd1dc5",
                              "6"="#d4ac0d","7"="#ca6f1e","8"="#797d7f","9"="#fa2c2c","10"="#3deb1e")
                   )

colors <- scale_color_manual(values = c("1"="#922b21","2"="#76448a","3"="#2471a3","4"="#17a589","5"="#cd1dc5",
                              "6"="#d4ac0d","7"="#ca6f1e","8"="#797d7f","9"="#fa2c2c","10"="#3deb1e")
                   )
```

Plot the points
```{r}
df_corn %>% 
  ggplot(aes(YOR,Yield)) +
  geom_point(aes(fill = StudyId),
             shape = 21,
             size = 4, alpha = 0.7) +
  fill_color +
  scale_x_continuous(limits = c(1940,2015))+
  scale_y_continuous(limits = c(0,21000)) +
  theme_bw()
```


## 2.2) Random Slope Model

The model:

$$y_{ij} = \beta_0 + \upsilon_i + \beta_1 x_j + \epsilon_{ij},$$

$$\upsilon_i \sim N(0,\sigma^2_{\upsilon}),$$

$$\epsilon_{ij} \sim N(0,\sigma^2_{\epsilon}).$$


$y_{ij}$ is the jth observation of corn grain yield in the ith study.

$\beta_0$ intercept.

$\upsilon_i$ effect of the ith randomly selected study over the intercept.

$\beta_1$ slope.

$x_j$ is the jth observation of the YOR.

$\epsilon_{ij}$ random error of the jth observation in the ith treatment.



### 2.2.2) Fitting the model

```{r}
options(scipen = 999)

m3_RI <- lmer(Yield ~ YOR + (1|StudyId),
     data = df_corn)

summary(m3_RI)
```


### 2.2.3) Obtain $\hat\upsilon_i$ and BLUPs

Borrow Strength (Sites 3 and 6)
```{r}
df_ups_BLUP <- as.data.frame(ranef(m3_RI)) %>%
           select(grp,condval) %>%
           rename(StudyId = grp, upsi.hat = condval) %>% 
           mutate(BLUP_int = m3_RI@beta[1]+upsi.hat,
                  slope = m3_RI@beta[2])

df_ups_BLUP

```


### 2.3.4) Plotting the reuslts

Prepare the data and make predictions
```{r}
df_list <- list()

for (i in c(as.factor(df_corn$StudyId))) {
  
  df_list[[i]] <-  filter(df_ups_BLUP,
         StudyId == i
         ) %>% 
    cross_join(.,
               data.frame(YOR = seq(min(df_corn$YOR),
                                    max(df_corn$YOR), 
                                    1))
               )
  
}

df_to_plot <- reduce(df_list,bind_rows) %>%
  mutate(Ey_hat = BLUP_int+slope*YOR)
  
```


Plot the data
```{r}
df_corn %>% 
  ggplot(aes(YOR,Yield)) +
  geom_point(aes(fill = StudyId),
             shape = 21,
             size = 4, alpha = 0.7) +
  fill_color +
  # Show the lines per study
  geom_line(data = df_to_plot,
            aes(YOR,Ey_hat,
                group = StudyId,
                color = StudyId),
            linetype = "twodash",
            linewidth = 1
            ) +
  colors+
  # Show the expected line
  geom_line(data = tibble(YOR = seq(min(df_corn$YOR),
                                        max(df_corn$YOR), 
                                        1),
                              Ey_hat = m3_RI@beta[1]+m3_RI@beta[2]*YOR
                              ),
            aes(YOR, Ey_hat),
            linetype = "solid",
            linewidth = 1
            ) +
  scale_x_continuous(limits = c(1940,2015))+
  scale_y_continuous(limits = c(0,21000)) +
  theme_bw()
```






