---
title: "Solutions for the first Assignments"
author: "Pedro Cisdeli"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
```
# 1. Data Structures
## 1.1 Create a Vector
Create a new chunk below this text. Then, create a vector called v1 with three elements, all of which are characters (enclosed in single ’ ’ or double ” quotes). Save it and print the values.
```{r 1.1}
v1 <- c("item1", "item2", "item3")
print(v1)
```
## 1.2 Convert vector to a factor
Create a vector v2 from v1, but convert it to a factor. Save and print the values. Change the order of the levels to any random order and print the values again.
```{r 1.2}
# Convert v1 to a factor
v2 <- factor(v1)
print(v2)

# Randomize the order of the levels
set.seed(123) # Set seed for reproducibility
random_levels <- sample(levels(v2)) # Randomize levels
v2 <- factor(v1, levels = random_levels)
print(v2)
```
## 1.3 Create a list
Create a list l1 that contains v1, v2, and an additional object of class data.frame containing the following information:


| Irrigation | Value |
|-----------:|------:|
| yes        | 200   |
|        yes |   200 |
|         no |     0 |
```{r 1.3}
# Create a data frame
df <- data.frame(
    Irrigation = c("yes", "yes", "no"),
    Value = c(200, 200, 0)
)

# Create a list containing v1, v2, and the data frame
l1 <- list(v1 = v1, v2 = v2, df = df)
print(l1)
```
# Tidyverse
## 2.1 Read the tillage data
Read the file doy_weather.xlsx from the “data” folder into a new data frame df.
```{r 2.1}
# Going back two directories, for some reason ./ is not working
df <- read_excel("../../Data/doy_weather.xlsx")
head(df, 5)
```
## 2.2 Check column types
Check the data types for all columns in the data frame. Some columns might be read as characters even though they represent numeric values. Convert any necessary columns to numeric.
```{r 2.2}
# We need to convert the following to numeric: TMIN TMAX RAIN SRAD
df <- df %>%
    mutate(
        TMIN = as.numeric(TMIN),
        TMAX = as.numeric(TMAX),
        RAIN = as.numeric(RAIN),
        SRAD = as.numeric(SRAD)
    )
head(df, 5)
```
## 2.3 Replace missing values
Some columns contain missing values coded as -99. Replace all instances of -99 with NA in numeric columns only. Save the result in a new data frame called df_clean.
```{r 2.3}
df_clean <- lapply(df, function(x) {
    x[x == -99] <- NA
    return(x)
})
df_clean <- as.data.frame(df_clean)
head(df_clean, 5)
```
## 2.4 Calculate mean Tmax
Calculate the mean of the TMAX column, ignoring missing values (NA). Save the result as mean_Tmax.
```{r 2.4}
df_clean <- df_clean %>% mutate(mean_Tmax = (mean(TMAX, na.rm = TRUE)))
head(df_clean, 5)
```
## 2.5 Compare temperatures
Create a new column Tmean, which is the average of TMAX and TMIN. Save the updated data frame as df_temp.
```{r 2.5}
df_temp <- df_clean %>% mutate(Tmean = (TMAX + TMIN) / 2)
head(df_temp, 5)
```
## 2.6 Day with the highest precipitation
Find the day with the highest precipitation (RAIN). Save the result as max_rain_day.
```{r 2.6}
max_rain_day <- df_temp %>% filter(RAIN == max(RAIN, na.rm = TRUE))
print(max_rain_day)
```
## 2.7 Subsetting data
Create a new data frame df_subset that includes only rows where precipitation (RAIN) is greater than 0.
```{r 2.7}
df_subset <- df_temp %>% filter(RAIN > 0)
head(df_subset, 5)
```
## 2.8 Sorting data
Sort the data frame df_clean by TMAX in descending order. Save the result as df_sorted.
```{r 2.8}
df_sorted <- df_temp %>% arrange(desc(TMAX))
head(df_sorted, 5)
```
## 2.9 Summary statistics
Calculate summary statistics (mean, median, standard deviation) for the TMAX column. Save the result as df_summary.
```{r 2.9}
df_summary <- df_temp %>% summarize(
    mean_Tmax = mean(TMAX, na.rm = TRUE),
    median_Tmax = median(TMAX, na.rm = TRUE),
    sd_Tmax = sd(TMAX, na.rm = TRUE)
)
head(df_summary)
```
## 2.10 Conditional Mutations with case_when
Create a new column Temperature_Category in df_clean, which classifies Tmean as:

“Low” if TMAX < 10
“Moderate” if 10 <= TMAX <= 25
“High” if TMAX > 25

Save the updated data frame as df_class selecting DOY, TMAX, and Category columns.
```{r 2.10}
df_class <- df_temp %>%
    mutate(Temperature_Category = case_when(
        TMAX < 10 ~ "Low",
        TMAX >= 10 & TMAX <= 25 ~ "Moderate",
        TMAX > 25 ~ "High"
    )) %>%
    select(DOY, TMAX, Temperature_Category)
head(df_class, 5)
```
